Enhancing diagnostic of stochastic mortality models leveraging contrast trees. An application on Italian data

### 2022-11-08 Manuscript Contrast Trees_R1.tex
Last version 8/11

%The literature relating to the mortality modeling and machine learning is growing. \color{blue} Citare le due scuole: ML come alternativa a canonical stochastic mortality models o come aggiuntivi \color{black} 
%to show how much better the machine learning model can do in terms of model fit and prediction accuracy

@@

%It is worth emphasizing that basically none of the above models is able to generalize on a wide-global dataset over age and time. Indeed, some models work adequately in a subset of countries, or even more on a subset of specific age-ranges. 
Despite models like the Lee-Carter and its variants having been widely used, becoming a benchmark for many newly proposed methodologies, they present several shortfalls, however. 
In this line, \cite{CBD08} tried to address the issue of what would be the best way to estimate mortality, exhibiting interesting criteria that a good mortality model should hold. They basically referred to good-practice guidelines such as the consistency with historical data and the long-term dynamics, biologically reasonable. Following this line of research, recent longevity literature stimulated the use of machine learning techniques in demographic research allowing the integration of stochastic models into a data-driven approach.

The significant reduction in the forecasting error reached by the application of machine learning techniques became particularly useful for both researchers and practitioners. The main contributions are from \cite{Deprez}, \cite{LevPiz} and \cite{LevNi}. The common idea behind all these works is to improve the fitting accuracy of canonical models using machine learning algorithms. In other words, to correct the mortality surface produced by standard stochastic mortality models. All of the proposed methods, calibrate a machine learning estimator used to adjust (and improve) the mortality rates estimated by the original mortality model. Those authors show that mortality modeling can benefit from machine learning that better captures patterns that traditional models do not identify. 
%Specifically, \cite{LevNi} proposed to extrapolate the machine learning estimator on the whole mortality surface using the two-dimensional P-splines, obtaining more accurate mortality projection with respect to the Lee-Carter model.

The need for new models that can understand mortality improvements more accurately than canonical tools is evident. 
This paper contributes to the literature on mortality evaluation by introducing an innovative approach based on machine learning techniques that demographers have not yet explored, contributing to the undervalued field of models assessment. 
Specifically, we provide an application of the method proposed by \cite{Friedman2020}, namely Contrast Trees, for evaluating the accuracy of the mortality estimates given by different models that are not treatable with standard validation methods (e.g. those based on the likelihood function). 
The key significance of the Contrast Trees in mortality evaluation is the supply of a unified approach allowing for assessing and comparing the accuracy of both standard mortality models and machine learning algorithms involved in mortality estimates. Moreover, in addition to evaluating the accuracy of the models, the Contrast Trees enables improving the performance of the models through a boosting procedure that reduces the inaccuracies. We use this methodology, namely Contrast Boosting, to improve the mortality rates estimates.

@@

We consider the Italian mortality data available in the Human Mortality Database (HMD) over the period 1950-2018. We refer to the male population aged 30-90, analyzing the age groups 30-60 and 61-90 separately to give greater evidence of the differences in mortality that characterizes the younger and older ages.
We split the data set into training set and test set according to the common rule 70\%-30\%. The dataset partition is obtained by using the dissimilarity-based compound selection proposed in \cite{Willett1999}.

@@

\subsection*{Mortality models}
In this section, we describe the four models to which the Contrast Trees methodology is applied to evaluate their quality of fit.
The first two models belong to the family of generalized age-period-cohort (GAPC) that are expressed in a regression framework to be suitable for applying Contrast Trees, which requires data organized in columns. The last two are well-known machine learning techniques also used for regression tasks.

In the following, we provide the models' specifications and brief descriptions.

\subsubsection*{Lee-Carter (LC) model}
The LC model\cite{LC92} assumes that:
\begin{equation} 
\label{eq:LC}
\log{\left(m_{x,t}\right)}=\alpha_{x}+\beta_{x} \kappa_{t}
\end{equation}
We further assume that deaths are independent Poisson distributed \cite{BDV2002}.
The age-specific parameter $\alpha_x$ provides the average age profile of mortality, the age-period term $\beta_{x}\cdot\kappa_t$ describes the mortality trends, with $\kappa_t$ the time index and $\beta_{x}$ modifying the effect of $\kappa_t$ across ages\footnote{
The model is subject to the following constraints on $\kappa_t$ and $\beta_x$: $\sum_{t \in \mathcal{T}}{\kappa_t}=0 \quad \sum_{x \in \mathcal{X}}{\beta_x}=1$.
Future mortality rates are obtained by modeling the time index $\kappa_{t}$ through an autoregressive integrated moving average (ARIMA) process. In general, a random walk with drift properly fits the data. 
%\begin{equation} 
%\kappa_{t} = \kappa_{t-1} + \delta + \epsilon_t, \quad \epsilon_t \sim N(0, \sigma^{2}_k)
%\end{equation}
%where $\delta$ is the drift parameter and $\epsilon_t$ are the error terms, normally distributed with null mean and variance $\sigma^{2}_k$.
}.
The LC model can be reformulated into a Generalized Nonlinear Models (GNM) framework according to the Poisson assumption for death counts.

\subsubsection*{Age-Period-Cohort (APC)}
We use the model's version reformulated into a Generalized Linear Models (GLM) framework \cite{Alai2014}:
\begin{equation} 
\label{eq:APC}
\log{\left(m_{x,t}\right)}=\beta_{0} +\beta_{1,x}+\beta_{2,t} +\beta_{3,x+t}
\end{equation}
Where the regression coefficients $\beta_{1,x}$, $\beta_{2,t}$, $\beta_{3,x+t}$ are the age trend, the period trend and the cohort trend.

@@

@@

These curves are shown in Fig. \ref{fig:lof} and Fig. \ref{fig:lof61} for the 30-60 and 61-90 age group, respectively. The left panels of these figures refer to the lack-of-fit curves obtained without applying the Contrast Boosting (Base CTs), while the right panels refer to the lack-of-fit curves obtained after applying Contrast Boosting to the output of the models (Boosted CTs).\\
For the 30-60 age group (Fig. \ref{fig:lof}), the APC model seems to best perform across all regions, as its discrepancy is consistently lower than that of other models. 

@@

@@

\subfloat[Base CTs]
{\includegraphics[width=0.51\textwidth]{lof_base}}
\subfloat[Boosted CTs]
{\includegraphics[width=0.51\textwidth]{lof_boosted}}
%:
\caption{Lack-of-fit contrast curves for APC (black), LC (blue), GBM (red), XGBM (orange) and XGBM prep (green). Age 30-60.}

@@

For the 61-90 age group (Fig. \ref{fig:lof61}), the GBM model seems by far the worst performing model. Although the application of the Contrast Boosting significantly reduces the discrepancy, the GBM remains less accurate than the other models. It should also be noted that the effect of Contrast Boosting in high-discrepancy regions for the other models is negligible, except for the APC.
\subfloat[Base CTs]
{\includegraphics[width=0.51\textwidth]{lof_61_base}}
\subfloat[Boosted CTs]
{\includegraphics[width=0.51\textwidth]{lof_61_boosted}}
\caption{Lack-of-fit contrast curves for APC (black), LC (blue), GBM (red), XGBM (orange) and XGBM prep (green). Age 61-90.}
\label{fig:lof61}

@@

However, the lack-of-fit curves provide more structured information than the average discrepancy, in particular, regarding how and how much $\bar{d}_m$ varies across the input space. This is clearer when comparing XGBM and APC models for the age group 61-90 in the base model: whereas the average discrepancy of these models differs for 2.6\% (0.002010 compared to 0.001960), looking at the lack-of-fit curve depicted in Fig. \ref{fig:lof61} the XGBM model (orange curve) results to perform better than the APC model (black curve) in the high discrepancy regions. 

@@

\begin{tabular}{|l|r|r|r|r|r|r|}

@@

\multirow{2}{*}{Model} & \multicolumn{3}{c|}{Age 30-60} & \multicolumn{3}{c|}{Age 61-90} \\
\cline{2-7}
			&	Base model	&	Boosted model	& \% Change &	Base model	&	Boosted model	& \% Change\\

@@

APC 			&	0.000150	(1)	&	0.000147 (1)	&	-2\%	&	0.002010 (3)	&	0.001740 (1)	&	-13\%\\
LC 			&	0.000229	(2)	&	0.000208 (3) 	&	-9\%	&	0.002130 (4)	&	0.002070 (4)	&	-3\%	\\
GBM 		&	0.000241	(3)	&	0.000167 (2)	&	-31\%&	0.006230 (5)	&	0.003400 (5)	&	-45\%\\
XGBM 		&	0.000283	(5)	&	0.000278 (5)	&	-2\%	&	0.001960 (1)	&	0.001950 (2)	&	-1\%	\\
XGBM prep 	&	0.000281	(4)	&	0.000272 (4)	&	-3\%	&	0.001970 (2) 	&	0.001970 (3)	&	0\%	\\

@@

@@

By comparing Table \ref{tab:rmse} showing the values of RMSE and Table \ref{tab:mape} showing the values of MAPE with Table \ref{tab:avecri} reporting the values of the average discrepancy, we note a greater convergence of the error measures in the boosted models rather than in the base models. This result is intuitively straightforward since the boosted models are obtained by just reducing the discrepancy measure. 
%
\begin{table}[ht]
\centering
\footnotesize
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\multirow{2}{*}{Model} & \multicolumn{3}{c|}{Age 30-60} & \multicolumn{3}{c|}{Age 61-90} \\
\cline{2-7}
			&	Base model	&	Boosted model	& \% Change &	Base model	&	Boosted model	& \% Change\\
\hline
APC 			&	0.000249	(1) 	& 0.000246	(1) 	& -1\% 	& 0.003796 	(4) & 0.003631	(3) & -4\%\\
LC 			&	0.000352	(2) 	& 0.000332	(3) 	& -6\% 	& 0.003779 	(3) & 0.003851	(4) & 2\%\\
GBM 		&	0.000465	(5) 	& 0.000294	(2) 	& -37\% 	& 0.012344 	(5) & 0.006111	(5) & -50\%\\
XGBM 		&	0.000379	(4)	& 0.000379	(5) 	& 0\% 	& 0.003070 	(2) & 0.003063	(2) & 0\%\\
XGBM prep 	&	0.000373	(3) 	& 0.000370	(4) 	& -1\% 	& 0.003047 	(1) & 0.003047	(1) & 0\%\\ 
\hline
\end{tabular}
\caption{Values of the RMSE}
\label{tab:rmse}
\end{table}
%
\begin{table}[ht]
\centering
\footnotesize
\begin{tabular}{|l|r|r|r|r|r|r|}
\hline
\multirow{2}{*}{Model} & \multicolumn{3}{c|}{Age 30-60} & \multicolumn{3}{c|}{Age 61-90} \\
\cline{2-7}
			&	Base model	&	Boosted model	& \% Change &	Base model	&	Boosted model	& \% Change\\
\hline
APC 			&	4.4\%	(1) & 4.3\%	(1) 	& -2\% 	& 3.7\%	(3) & 3.1\%	(1) & -18\%\\
LC 			&	6.9\%	(3) & 5.9\%	(3) 	& -15\% 	& 4.7\%	(4) & 4.6\%	(4) & -3\%\\
GBM			&	10.3\%	(5) & 5.7\%	(2) 	& -44\% 	& 17.7\%	(5) & 11.6\%	(5) & -35\%\\
XGBM		&	7.2\%	(4) & 6.9\%	(5) 	& -5\% 	& 3.6\%	(2) & 3.5\%	(3) & -1\%\\
XGBM prep	&	6.9\%	(2) & 6.5\%	(4) 	& -6\% 	& 3.5\%	(1) & 3.5\%	(2) & 0\%\\ 
\hline
\end{tabular}
\caption{Values of the MAPE}
\label{tab:mape}
\end{table}
%
Results of the CTs, comparing the estimates of the models with the real mortality rates, provide the discrepancy measure for each region detected by the CTs. %\hl{(Appendix)}. 
These regions can easily be identified and, if possible, interpreted, providing a further explanation to the model performances. Moreover, high-discrepancy regions can be used for assessing whether or where a model should be trusted or not. We show the heatmap of the error regions in Fig. \ref{fig:CT} and Fig. \ref{fig:CT2} for ages 30-60 and 61-90, respectively. Low discrepancy regions are painted in green, while high discrepancy regions are painted in red. Purple is used, for the sake of image readability, for regions whose discrepancy value exceeds 6e-04 and 0.008 respectively.

@@

\begin{figure}[H]
\centering
\subfloat[APC (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{apc.jpeg}}
{\includegraphics[width=0.25\textwidth]{apc_b.jpeg}}\\
\subfloat[LC(left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{lc.jpeg}}
{\includegraphics[width=0.25\textwidth]{lc_b.jpeg}}\\
\subfloat[GBM (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{gbm.jpeg}}
{\includegraphics[width=0.25\textwidth]{gbm_b.jpeg}}\\
\subfloat[XGBM (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{xgbm.jpeg}}
{\includegraphics[width=0.25\textwidth]{xgbm_b.jpeg}}\\
\subfloat[XGM prep (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{xgbm_preproc.jpeg}}
{\includegraphics[width=0.25\textwidth]{xgbm_preproc_b.jpeg}}\\
\caption{CTs regions, age 30-60.}
\label{fig:CT}
\end{figure}
%
\begin{figure}[H]
\centering
\subfloat[APC (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{apc_61.jpeg}}
{\includegraphics[width=0.25\textwidth]{apc_61_b.jpeg}}\\
\subfloat[LC (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{lc_61.jpeg}}
{\includegraphics[width=0.25\textwidth]{lc_61_b.jpeg}}\\
\subfloat[GBM (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{gbm_61.jpeg}}
{\includegraphics[width=0.25\textwidth]{gbm_61_b.jpeg}}\\
\subfloat[XGBM (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{xgbm_61.jpeg}}
{\includegraphics[width=0.25\textwidth]{xgbm_61_b.jpeg}}\\
\subfloat[XGM prep (left: base; right: boosted)]
{\includegraphics[width=0.25\textwidth]{xgbm_preproc_61.jpeg}}
{\includegraphics[width=0.25\textwidth]{xgbm_preproc_61_b.jpeg}}\\
\caption{CTs regions, age 61-90.}
\label{fig:CT2}
\end{figure}
%\subsubsection*{Third-level section}
 %Topical subheadings are allowed.

@@

@@

@@

>>>>

@@

Despite models like the Lee-Carter and its variants having been widely used, becoming a benchmark for many newly proposed methodologies, they present several shortfalls, however. 
In this line, \cite{CBD08} tried to address the issue of what would be the best way to estimate mortality, exhibiting interesting criteria that a good mortality model should hold. They basically referred to good-practice guidelines such as the consistency with historical data and the long-term dynamics, biologically reasonable. Following this line of research, recent longevity literature stimulated the use of machine learning techniques in demographic research allowing the integration of stochastic models into a data-driven approach.

The significant reduction in the forecasting error reached by the application of machine learning techniques became particularly useful for both researchers and practitioners. The main contributions are from \cite{Deprez}, \cite{LevPiz} and \cite{LevNi}. The common idea behind all these works is to improve the fitting accuracy of canonical models using machine learning algorithms. In other words, to correct the mortality surface produced by standard stochastic mortality models. All of the proposed methods, calibrate a machine learning estimator used to adjust (and improve) the mortality rates estimated by the original mortality model. Those authors show that mortality modeling can benefit from machine learning that better captures patterns that traditional models do not identify.

The need for new models that can understand mortality improvements more accurately than canonical tools is evident. 
This paper contributes to the literature on mortality evaluation by introducing an innovative approach based on machine learning techniques that demographers have not yet explored, contributing to the undervalued field of models assessment. 
\color{blue}
We provide an application of the method proposed by \cite{Friedman2020}, namely Contrast Trees, which helps evaluate the accuracy of the mortality estimates (fitted mortality rates) given by models that are not treatable with model selection criteria based on the likelihood function. Therefore, this technique provides a unified framework for assessing and comparing the goodness-of-fit to historical data of traditional mortality models with machine learning algorithms.
\color{black}
Moreover, in addition to evaluating the accuracy of the models, the Contrast Trees enables improving the performance of the models through a boosting procedure that reduces the inaccuracies. We use this methodology, namely Contrast Boosting, to improve \color{blue}the fitting of historical mortality data.
To summarize, through this new technique based on Contrast Trees, we aim to find the best model that fits historical mortality rates by grasping and detecting the inaccuracies of any model and boosting its predictive power. \color{black}

@@

We consider the Italian mortality data available in the Human Mortality Database (HMD) over the period 1950-2018. We refer to the male population aged \color{blue}0-90, analyzing the age groups 0-29, 30-60, and 61-90 separately to provide further evidence of the differences in mortality that characterizes the younger ages, the adult ages, and the older ages. 
We split the data set into training set (fitting period) and test set (out-of-sample forecast period) according to the common rule 70\%-30\%. We will calculate the in-sample errors using data of the training set, and the out-of-sample errors using data of the test set (not used to build the training predictor).
\color{black}
The dataset partition is obtained by using the dissimilarity-based compound selection proposed in \cite{Willett1999}.

@@

\subsection*{Mortality models}
In this section, we describe the four models to which the Contrast Trees methodology is applied to evaluate their quality of fit.
The first two models belong to the family of generalized age-period-cohort (GAPC) that are expressed in a regression framework to be suitable for applying Contrast Trees, which requires data organized in columns. The last two are well-known machine learning techniques also used for regression tasks.\\
In the following, we provide the models' specifications and brief descriptions.

\subsubsection*{Lee-Carter (LC) model}
The original LC model\cite{LC92} assumes that:
\begin{equation} 
\label{eq:LC}
\log{\left(m_{x,t}\right)}=\alpha_{x}+\beta_{x} \kappa_{t}
\end{equation}
The age-specific parameter $\alpha_x$ provides the average age profile of mortality, the age-period term $\beta_{x}\cdot\kappa_t$ describes the mortality trends, with $\kappa_t$ the time index and $\beta_{x}$ modifying the effect of $\kappa_t$ across ages. The model is subject to the following constraints on $\kappa_t$ and $\beta_x$: $\sum_{t \in \mathcal{T}}{\kappa_t}=0 \quad \sum_{x \in \mathcal{X}}{\beta_x}=1$.
Future mortality rates are obtained by modeling the time index $\kappa_{t}$ through an autoregressive integrated moving average (ARIMA) process. In general, a random walk with drift properly fits the data. \\
%\begin{equation} 
%\kappa_{t} = \kappa_{t-1} + \delta + \epsilon_t, \quad \epsilon_t \sim N(0, \sigma^{2}_k)
%\end{equation}
%where $\delta$ is the drift parameter and $\epsilon_t$ are the error terms, normally distributed with null mean and variance $\sigma^{2}_k$.
\color{blue}The LC model can be reformulated into a Generalized Non-linear Models (GNM) framework following the approach proposed by \cite{BDV2002} assuming that deaths are independent Poisson distributed. The authors use a GNM and apply the maximum likelihood method to fit the model to historical data. Under this specification, the LC model can be seen as a regression model where mortality rates are the target variable, predicted using features (age and time) \cite{RichmanWuthrich}. \color{black}

\subsubsection*{Age-Period-Cohort (APC)}
We use the model's version reformulated into a Generalized Linear Models (GLM) framework \cite{Alai2014}:
\begin{equation} 
\label{eq:APC}
\color{blue} \log{\left(m_{x,t}\right)}=\beta_{0} +\beta_{1,x}+\beta_{2,t} +\beta_{3,t-x}
\end{equation}
Where the regression coefficients $\beta_{1,x}$, $\beta_{2,t}$, \color{blue}$\beta_{3,t-x}$\color{black} are the age trend, the period trend and the cohort trend \color{blue}($t-x$ represents the year of birth)\color{black}.

@@

\color{blue}
\subsection*{Traditional diagnostic tools}
In the following, we briefly mention some traditional diagnostic tools that are often used in the literature to assess the goodness-of-fit of a mortality model. 
\begin{itemize}
\item Analysis of mortality residuals (or standardized mortality residuals) calculated as the difference between the crude estimate of mortality rate by age and year based on observed data and the corresponding estimated mortality rate using a specified mortality model. For example, \cite{Cairns2010} verified that they are consistent with the hypothesis of i.i.d. N(0,1) and have zero correlation both across adjacent ages and across adjacent years.
\item Proportion of variance explained ($R^2$) by the model or the parameters of the model (see, e.g., \cite{Bongaarts2005})
\item Model selection criteria that penalize the log-likelihood with the increase in number of parameters: Akaike Information Criterion (AIC), Schwarz-Bayes Criterion (SBC) (or Bayes Information Criterion (BIC)) and Likelihood-ratio test (LRT) (\cite{Li2009}). Note that in this case the evaluation of the goodness-of-fit is given on the basis of the log-likelihood.
\item Checking for the absence of autocorrelation in the residuals of the model by the Portmanteau test (see, e.g., \cite{Torri2011}).
\end{itemize}
\color{black}

@@

\color{blue} These curves are shown in Fig. \ref{fig:lof} for the three different ages groups analyzed. The panels (a)-(c) of these figures refer to the lack-of-fit curves obtained without applying the Contrast Boosting (Base models), while panels (d)-(f) refer to the lack-of-fit curves obtained after applying Contrast Boosting to the output of the models (Boosted models).\\
For the 0-29 age group (Fig. \ref{fig:lof}, panel (a) and (d)), ADD COMMENTS \color{black}
For the 30-60 age group (Fig. \ref{fig:lof}, panel (b) and (e)), the APC model seems to best perform across all regions, as its discrepancy is consistently lower than that of other models. 

@@

For the 61-90 age group (Fig. \ref{fig:lof}, panel (c) and (f)), the GBM model seems by far the worst performing model. Although the application of the Contrast Boosting significantly reduces the discrepancy, the GBM remains less accurate than the other models. It should also be noted that the effect of Contrast Boosting in high-discrepancy regions for the other models is negligible, except for the APC.

@@

\subfloat[Base model - Age 0-29]
{\includegraphics[width=0.3\textwidth]{lof_0}}
\subfloat[Base model - Age 30-60]
{\includegraphics[width=0.3\textwidth]{lof}}
\subfloat[Base model - Age 61-90]
{\includegraphics[width=0.3\textwidth]{lof_61_base}}\\
\subfloat[Boosted model - Age 0-29]
{\includegraphics[width=0.3\textwidth]{lof_0_b}}
\subfloat[Boosted model - Age 30-60]
{\includegraphics[width=0.3\textwidth]{lof_b}}
\subfloat[Boosted model - Age 61-90]
{\includegraphics[width=0.3\textwidth]{lof_61_boosted}}
\caption{Lack-of-fit contrast curves for APC (black), LC (blue), GBM (red), XGBM (orange) and XGBM prep (green).}

@@

@@

However, the lack-of-fit curves provide more structured information than the average discrepancy, in particular, regarding how and how much $\bar{d}_m$ varies across the input space. This is clearer when comparing XGBM and APC models for the age group 61-90 in the base model: whereas the average discrepancy of these models differs for 2.6\% (0.002010 compared to 0.001960), looking at the lack-of-fit curve depicted in Fig. \ref{fig:lof} the XGBM model (orange curve) results to perform better than the APC model (black curve) in the high discrepancy regions. 

@@

\begin{tabular}{|l|r|r|r|r|r|r|r|r|r|}

@@

\multirow{2}{*}{Model} & \multicolumn{3}{c|}{Age 0-29} & \multicolumn{3}{c|}{Age 30-60} & \multicolumn{3}{c|}{Age 61-90} \\
\cline{2-10}
			&	Base 	&	Boosted 	& \% Change &	Base 	&	Boosted 	& \% Change &	Base 	&	Boosted	& \% Change\\

@@

APC 	&	0.000410	&	0.000409	&	0\%	&	0.000145	&	0.000141	&	-3\%	&	0.002142	&	0.001948	&	-9\%	\\
LC 	&	0.000571	&	0.000568	&	0\%	&	0.000231	&	0.000232	&	0\%	&	0.002395	&	0.002314	&	-3\%	\\
GBM 	&	0.000473	&	0.000459	&	-3\%	&	0.000320	&	0.000225	&	-30\%	&	0.006525	&	0.003238	&	-50\%	\\
XGBM 	&	0.000417	&	0.000415	&	0\%	&	0.000240	&	0.000233	&	-3\%	&	0.001916	&	0.001940	&	1\%	\\
XGBM prep 	&	0.000500	&	0.000493	&	-1\%	&	0.000268	&	0.000265	&	-1\%	&	0.002003	&	0.002005	&	0\%	\\
%APC 		&	0.000150	(1)	&	0.000147 (1)	&	-2\%	&	0.002010 (3)	&	0.001740 (1)	&	-13\%\\
%LC 		&	0.000229	(2)	&	0.000208 (3) 	&	-9\%	&	0.002130 (4)	&	0.002070 (4)	&	-3\%	\\
%GBM 		&	0.000241	(3)	&	0.000167 (2)	&	-31\%&	0.006230 (5)	&	0.003400 (5)	&	-45\%\\
%XGBM 		&	0.000283	(5)	&	0.000278 (5)	&	-2\%	&	0.001960 (1)	&	0.001950 (2)	&	-1\%	\\
%XGBM prep 	&	0.000281	(4)	&	0.000272 (4)	&	-3\%	&	0.001970 (2) 	&	0.001970 (3)	&	0\%	\\

@@

\end{table}\\

@@

By comparing Table \ref{tab:errors} showing the values of RMSE and MAPE with Table \ref{tab:avecri} reporting the values of the average discrepancy, we note a greater convergence of the error measures in the boosted models rather than in the base models. This result is intuitively straightforward since the boosted models are obtained by just reducing the discrepancy measure. 
\begin{tabular}{|l|l|r|r|r|r|r|r|r|r|r|}
\multirow{2}{*}{Error} & \multirow{2}{*}{Model} & \multicolumn{3}{c|}{Age 0-29} & \multicolumn{3}{c|}{Age 30-60} & \multicolumn{3}{c|}{Age 61-90} \\
\cline{3-11}
&			&	Base 	&	Boosted 	& \% Change &	Base 	&	Boosted 	& \% Change &	Base 	&	Boosted	& \% Change\\
\multirow{5}{*}{RMSE}  & APC 	&	0.002040	&	0.002039	&	0\%	&	0.000264	&	0.000263	&	0\%	&	0.004260	&	0.004139	&	-3\%	\\
& LC 	&	0.003471	&	0.003471	&	0\%	&	0.000491	&	0.000496	&	1\%	&	0.004258	&	0.004363	&	2\%	\\
& GBM 	&	0.001648	&	0.001647	&	0\%	&	0.000640	&	0.000455	&	-29\%	&	0.012248	&	0.005439	&	-56\%	\\
& XGBM 	&	0.001517	&	0.001515	&	0\%	&	0.000342	&	0.000338	&	-1\%	&	0.003260	&	0.003278	&	1\%	\\
& XGBM prep 	&	0.001939	&	0.001935	&	0\%	&	0.000391	&	0.000386	&	-1\%	&	0.003339	&	0.003345	&	0\%	\\
%APC 		&	0.000249	(1) 	& 0.000246	(1) 	& -1\% 	& 0.003796 	(4) & 0.003631	(3) & -4\%\\
%LC 		&	0.000352	(2) 	& 0.000332	(3) 	& -6\% 	& 0.003779 	(3) & 0.003851	(4) & 2\%\\
%GBM 		&	0.000465	(5) 	& 0.000294	(2) 	& -37\% 	& 0.012344 	(5) & 0.006111	(5) & -50\%\\
%XGBM 		&	0.000379	(4)	& 0.000379	(5) 	& 0\% 	& 0.003070 	(2) & 0.003063	(2) & 0\%\\
%XGBM prep 	&	0.000373	(3) 	& 0.000370	(4) 	& -1\% 	& 0.003047 	(1) & 0.003047	(1) & 0\%\\ 
%\hline
%\end{tabular}
%\caption{Values of the RMSE}
%\label{tab:rmse}
%\end{table}
%%
%\begin{table}[ht]
%\centering
%\footnotesize
%\begin{tabular}{|l|l|r|r|r|r|r|r|r|r|r|}
%\hline
%\multirow{2}{*}{Error} & \multirow{2}{*}{Model} & \multicolumn{3}{c|}{Age 0-29} & \multicolumn{3}{c|}{Age 30-60} & \multicolumn{3}{c|}{Age 61-90} \\
%\cline{3-11}
%&			&	Base 	&	Boosted 	& \% Change &	Base 	&	Boosted 	& \% Change &	Base 	&	Boosted	& \% Change\\
\multirow{5}{*}{MAPE}  & APC 			&	14.7\%	&	14.5\%	&	-1\%		&	4.5\%	&	4.3\%	&	-3\%		&	3.9\%	&	3.4\%	&	-14\% \\
& LC 			&	14.2\%	&	13.8\%	&	-3\%		&	7.2\%	&	7.1\%	&	-1\%		&	4.9\%	&	4.9\%	&	0\%	\\
& GBM 		&	23.4\%	&	18.8\%	&	-20\%	&	13.0\%	&	7.6\%	&	-41\%	&	18.3\%	&	9.2\%	&	-50\% \\
& XGBM 		&	15.9\%	&	15.3\%	&	-3\%		&	6.9\%	&	6.2\%	&	-10\%	&	3.7\%	&	3.8\%	&	2\%	\\
& XGBM prep 	&	20.0\%	&	18.2\%	&	-9\%		&	7.3\%	&	7.2\%	&	-1\%		&	3.6\%	&	3.6\%	&	0\%	\\
%APC 		&	4.4\%	(1) & 4.3\%	(1) 	& -2\% 	& 3.7\%	(3) & 3.1\%	(1) & -18\%\\
%LC 		&	6.9\%	(3) & 5.9\%	(3) 	& -15\% 	& 4.7\%	(4) & 4.6\%	(4) & -3\%\\
%GBM		&	10.3\%	(5) & 5.7\%	(2) 	& -44\% 	& 17.7\%	(5) & 11.6\%	(5) & -35\%\\
%XGBM		&	7.2\%	(4) & 6.9\%	(5) 	& -5\% 	& 3.6\%	(2) & 3.5\%	(3) & -1\%\\
%XGBM prep	&	6.9\%	(2) & 6.5\%	(4) 	& -6\% 	& 3.5\%	(1) & 3.5\%	(2) & 0\%\\ 
\caption{Values of the RMSE and MAPE}
\label{tab:errors}
\color{blue}
We also calculate average discrepancy, RMSE, and MAPE on the logarithm of the central death rates (Table \ref{tab:avecri_log}, \ref{tab:errors_log}. 
These measures assign a relatively large weight to errors at young ages, while error measures calculated on the central death rates assign a large weight to errors at older ages.
\color{black} 
Results of the CTs, comparing the estimates of the models with the real mortality rates, provide the discrepancy measure for each region detected by the CTs. %\hl{(Appendix)}. 
These regions can easily be identified and, if possible, interpreted, providing a further explanation to the model performances. Moreover, high-discrepancy regions can be used for assessing whether or where a model should be trusted or not. \color{blue}We show the heatmap of the error regions in Fig. \ref{fig:CT} and Fig. \ref{fig:CT2} for the base model and the boosted one, respectively. \color{black} Low discrepancy regions are painted in green, while high discrepancy regions are painted in red. Purple is used, for the sake of image readability, for regions whose discrepancy value exceeds 6e-04 and 0.008 respectively.

@@

\begin{figure}[H]
\centering
\subfloat[APC (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{apc_0.jpeg}}
{\includegraphics[width=0.27\textwidth]{apc.jpeg}}
{\includegraphics[width=0.27\textwidth]{apc_61.jpeg}}
\subfloat[LC (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{lc_0.jpeg}}
{\includegraphics[width=0.27\textwidth]{lc.jpeg}}
{\includegraphics[width=0.27\textwidth]{lc_61.jpeg}}
\subfloat[GBM (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{gbm_0.jpeg}}
{\includegraphics[width=0.27\textwidth]{gbm.jpeg}}
{\includegraphics[width=0.27\textwidth]{gbm_61.jpeg}}
\subfloat[XGBM (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{xgbm_0.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_61.jpeg}}

\subfloat[XGM prep (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{xgbm_preproc_0.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_preproc.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_preproc_61.jpeg}}
\caption{CTs regions, Base model.}
\label{fig:CT}
\end{figure}
%
\begin{figure}[H]
\centering
\subfloat[APC (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{apc_0_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{apc_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{apc_61_b.jpeg}}
\subfloat[LC (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{lc_0_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{lc_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{lc_61_b.jpeg}}
\subfloat[GBM (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{gbm_0_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{gbm_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{gbm_61_b.jpeg}}
\subfloat[XGBM (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{xgbm_0_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_61_b.jpeg}}
\subfloat[XGM prep (left: age 0-29; centre: age 30-60 right: age 61-90)]
{\includegraphics[width=0.27\textwidth]{xgbm_preproc_0_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_preproc_b.jpeg}}
{\includegraphics[width=0.27\textwidth]{xgbm_preproc_61_b.jpeg}}
\caption{CTs regions, Boosted model.}
\label{fig:CT2}
\end{figure}
%\subsubsection*{Third-level section}
 %Topical subheadings are allowed.

@@

\bibitem[Bongaarts(2005)]{Bongaarts2005}
Bongaarts, J. (2005). Long-Range Trends in Adult Mortality: Models and Projection Methods. Demography, 42(1): 23-49.

@@

\bibitem[Cairns et al.(2010)]{Cairns2010}
Cairns, A.J.G., Blake, D., Dowd, K., Coughlan, G.D., Epstein, D., Khalaf-Allah, M. (2010). Evaluating the goodness of fit of stochastic mortality models. Insurance: Mathematics and Economics, 47(3): 255-265.

@@

\bibitem[Li et al.(2009)]{Li2009}
Li, J. S.H.,  Hardy, M. R., Tan, K. S. (2009). Uncertainty in mortality forecasting: an extension to the classical Lee-Carter approach. Astin Bulletin 39(1), 137-164.
